---
title: "assignment3_CM"
author: "Christopher Mallon"
date: "22/12/2021"
output: html_document
---

All three problems below are based on the same data. The data in data(foxes)
are 116 foxes from 30 different urban groups in England. These foxes are like
street gangs. Group size varies from 2 to 8 individuals. Each group main-
tains its own (almost exclusive) urban territory. Some territories are larger
than others. The area variable encodes this information. Some territories
also have more avgfood than others. We want to model the weight of each
fox. For the problems below, assume this DAG:

area -> avgfood -> groupsize -> weight     /\   /\   Todd Vargo
area -> avgfood -> weight                 //\\_//\\     ____
                                          \_     _/    /   /
                                           / * * \    /^^^]
                                           \_\O/_/    [   ]
                                            /   \_    [   /
                                            \     \_  /  /
                                             [ [ /  \/ _/
                                            _[ [ \  /_/
                                            
```{r Setup}
rm(list = ls())
gc()
library(rethinking)
data(foxes)
d <- foxes
```

# Useful stats
Average weight of a fox is `r mean(d$weight)`, max is `r max(d$weight)`, sd is `r sd(d$weight)`.
<!-- avg: 4.52, max: 7.55, sd: 1.18 -->

# Problem 1
Use a model to infer the total causal influence of area on weight . Would
increasing the area available to each fox make it heavier (healthier)? You
might want to standardize the variables. Regardless, use prior predictive
simulation to show that your model’s prior predictions stay within the pos-
sible outcome range.

## Assess overall relationship
```{r area-weight relationship}
plot(d$weight~d$area)
```

```{r So let's try to infer the total causal influence}
# standardize vars
d$A <- (d$area - mean(d$area))/sd(d$area)
d$W <- (d$weight - mean(d$weight)) / sd(d$weight)

# W_i~Normal(μ_i,σ)
# μ_i = α + β_A*A_i
# α ~ Normal(4.5, 2) # guessing avg. fox weight is 4.5, sd 2 (I cheated)
# β_A ~ Normal(0, 0.5) # guessing slope of 0, up to 0.5 (approx 22.5 degrees)
# σ ~ Exponential(1) # Not sure what this means, but it's recommended

area_weight_model <- quap(
  alist(
    W ~ dnorm(mu_i, sigma), 
    mu_i <- alpha + beta_A*A, # Does 'A' link to our d$A?
    alpha ~ dnorm(0, 0.5), # average, w. sd 1 - so far I dislike standardization
    beta_A ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d
)

set.seed(1)
prior <- extract.prior(area_weight_model)
mu <- link(area_weight_model, post = prior, data = list(A=c(-2,2)))
plot( NULL , xlim=c(-3,3) , ylim=c(-3,3), xlab="area" , ylab="weight" )
for ( i in 1:50 ) lines( c(-2,2) , mu[i,] , col=col.alpha("black",0.4) )
```

I think the intention here is to show that there's no relationship. My quesion is: does the above do that?

```{r}
# percentile interval of mean
A_seq = seq(from=-2.5, to=2.5, length.out=50)
mu <- link(area_weight_model, data = list(A=A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# plot
plot(W~A, data = d, col = rangi2)
lines(A_seq, mu.mean, lwd=2)
shade(mu.PI, A_seq)
```


# Problem 2
Now infer the causal impact of adding food to a territory. Would this
make foxes heavier? Which covariates do you need to adjust for to estimate
the total causal influence of food?

```{r}
par(mfrow=c(1,2))
plot(d$avgfood~d$area)
plot(d$weight~d$avgfood)
```

It seems that area size and avg food are correlated. Avgfood and weight do not
seem to correlate.

```{r}
d$AF <- (d$avgfood - mean(d$avgfood)) / sd(d$avgfood)

area_food_weight <- quap(
  alist(W ~ dnorm(mu_i, sigma),
        mu_i <- alpha + B1*A+B2*AF,
        alpha ~ dnorm(0,0.2),
        B1~dnorm(0, 0.5),
        B2~dnorm(0, 0.5),
        sigma~dexp(1)
        ),
  data=d
)

precis(area_food_weight)

```

```{r}
plot(coeftab(area_food_weight, area_weight_model), par=c("B1")) # why doesn't this work?
```


# Problem 3
Now infer the causal impact of group size. Which covariates do you need
to adjust for? Looking at the posterior distribution of the resulting model,
what do you think explains these data? That is, can you explain the estimates
for all three problems? How do they go together?

```{r}
d$GS <- (d$groupsize - mean(d$groupsize)) / sd(d$groupsize)

area_food_group_weight <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- alpha + B1*A + B2*AF + B3*GS,
    alpha ~ dnorm(0, 1),
    B1 ~ dnorm(0, 0.5),
    B2 ~ dnorm(0, 0.5),
    B3 ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d
)
precis(area_food_group_weight)

```

So, as group size increases, weight decreases? That makes some intuitive sense.
Additionally, now both B1 and B2 have positive slopes, when they didn't before.
So maybe factoring in group size can show that both area and food are overall
positive? And that group size confounds that by being negative?