---
title: "assignment5_CM"
author: "Christopher Mallon"
date: "29/04/2022"
output: html_document
---


# 1. 
Consider the data(Wines2012) data table. These data are expert ratings
of 20 different French and American wines by 9 different French and Amer-
ican judges. Your goal is to model score , the subjective rating assigned by
each judge to each wine. I recommend standardizing it.
In this first problem, consider only variation among judges and wines.
Construct index variables of judge and wine and then use these index vari-
ables to construct a linear regression model. Justify your priors. You should
end up with 9 judge parameters and 20 wine parameters. Use ulam instead of
quap to build this model, and be sure to check the chains for convergence. If
you’d rather build the model directly in Stan or PyMC3, go ahead. I just want
you to use Hamiltonian Monte Carlo instead of quadratic approximation.
How do you interpret the variation among individual judges and indi-
vidual wines? Do you notice any patterns, just by plotting the differences?
Which judges gave the highest/lowest ratings? Which wines were rated worst/
best on average?

```{r}
library(rethinking)
data("Wines2012")
d <- Wines2012
d$j_index <- as.integer(as.factor(d$judge))
d$w_index <- as.integer(as.factor(d$wine))
d$score_std <- (d$score - min(d$score))/(max(d$score) - min(d$score))

d_slim <- list(
j_index = d$j_index,
w_index = d$w_index,
score = d$score_std
)

hm5.1 <- ulam(
  alist(
    score ~ dnorm(mu, sigma),
    mu <- a[j_index] + b[w_index], # How were we supposed to know to do this? Is this even a lm anymore? WHich one is slope?
    a[j_index] ~ dnorm(0.5, 0.2), # because a judge could like or dislike a wine
    b[w_index] ~ dnorm(0.5, 0.2),
    sigma ~ dexp(1)
  ), data = d_slim, chains = 4
)

precis(hm5.1, depth = 2)

```

```{r}
traceplot(hm5.1)
```

In spite of some interesting Rhat values (>1, uh oh) the traceplots look okay. Looking at the results, 
it does seem like there were some wines that were generally less preferred than others. It also
looks like judges with lower coefficients were globally harsher.

The harsher judges were Jean-M Cardebat (4), and Robert Hodgson (8). The least favourite wine was I2(18), followed by C2(6).


# 2. 
Now consider three features of the wines and judges:
(1) flight : Whether the wine is red or white.
(2) wine.amer : Indicator variable for American wines.
(3) judge.amer : Indicator variable for American judges.
Use indicator or index variables to model the influence of these features on
the scores. Omit the individual judge and wine index variables from Problem
1. Do not include interaction effects yet. Again use ulam , justify your priors,
and be sure to check the chains. What do you conclude about the differences
among the wines and judges? Try to relate the results to the inferences in
Problem 1.

```{r}
data("Wines2012")
d <- Wines2012
d$f_index <- as.integer(as.factor(d$flight))
d$wa_index <- as.integer(as.factor(d$wine.amer))
d$ja_index <- as.integer(as.factor(d$judge.amer))
d$score_std <- (d$score - min(d$score))/(max(d$score) - min(d$score))

d_slim <- list(
  f_index = d$f_index,
  ja_index = d$ja_index,
  wa_index = d$wa_index,
  score = d$score_std
)

hm5.2 <- ulam(
  alist(
    score ~ dnorm(mu, sigma),
    mu <- a[f_index] + b[ja_index] + c[wa_index], # the solution says that htis is wrong. Why? Help?
    a[f_index] ~ dnorm(0, 0.2), # set priors to 0 b/c we don't know if the influence will be positive or negative.
    b[ja_index] ~ dnorm(0, 0.2),
    c[wa_index] ~ dnorm(0, 0.2),
    sigma ~ dexp(1)
  ), data = d_slim, chains = 4
)

precis(hm5.2, depth = 2)
```
Let's just see how it looks first.

```{r}
pairs(hm5.2@stanfit)
```
Looks like the index vars are correlated w. each other, but otherwise looks like an okay spread.

```{r}
traceplot(hm5.2)
```

Those are some nice fuzzy caterpillars.
So thoughts:
Not a lot of difference between the provenance, the flight, or the judge nationality. Of course, this is only when each one is looked at separately.

# 3. 
Now consider two-way interactions among the three features. You should
end up with three different interaction terms in your model. These will be
easier to build, if you use indicator variables. Again use ulam , justify your
priors, and be sure to check the chains. Explain what each interaction means.
Be sure to interpret the model’s predictions on the outcome scale ( mu , the
expected score), not on the scale of individual parameters. You can use link
to help with this, or just use your knowledge of the linear model instead.
What do you conclude about the features and the scores? Can you relate
the results of your model(s) to the individual judge and wine inferences from
Problem 1?

```{r}
data("Wines2012")
d <- Wines2012
d$f <- as.integer(as.factor(d$flight))
d$wa <- as.integer(as.factor(d$wine.amer))
d$ja <- as.integer(as.factor(d$judge.amer))
d$score_std <- (d$score - min(d$score))/(max(d$score) - min(d$score))

d_slim <- list(
  f = d$f,
  ja = d$ja,
  wa = d$wa,
  score = d$score_std
)

hm5.3 <- ulam(
  alist(
    score ~ dnorm(mu, sigma),
    mu <- bf*f + bj*ja + bw*wa + bfj*f*ja + bfw*f*wa + bjw*ja*wa,
    bf ~ dnorm(0, 0.2), # set priors to 0 b/c we don't know if the influence will be positive or negative.
    bj ~ dnorm(0, 0.2),
    bw ~ dnorm(0, 0.2),
    bfj~ dnorm(0, 0.2),
    bfw ~ dnorm(0, 0.2),
    bjw ~ dnorm(0, 0.2),
    sigma ~ dexp(1)
  ), data = d_slim, chains = 4
)

precis(hm5.3, depth = 2)
```

Neat. So in this final model, it does appear that ther is a substantial judge effect. What could that indicate?

I'm also seeing a negative effect having to do with the interaciton between flight and judge, and also between judge.american and wine.american. I'm a bit surprised because the taste test was blind - why would these have interaction effects?